A Plasma Torus Around a Young Low-Mass Star

### 2025-03-26 paper/ms.tex
update ms

M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water \cite{NAP26141}.
Community investment with JWST has proceeded accordingly
\cite{Redfield2024,TRAPPIST1JWSTCommunityInitiative2024}.  How an M
dwarf's evolution influences its planets---especially the retention of
their atmospheres--- has in turn become a major theme in
exoplanet and stellar astrophysics.  Previous work has established
that most M dwarfs host close-in planets \cite{Dressing2015} that are
on average subject to long circumstellar disk lifetimes
\cite{Ribas2015}, to large doses of UV radiation \cite{France2013},
and to a high incidence of flares and coronal mass ejections
\cite{Gunther2020}.  However, despite extensive work in these areas,
the plasma and magnetospheric environments that bathe young, close-in
exoplanets remain challenging to quantify.
Understanding these environments is crucial because they directly
impact atmospheric retention and habitability of close-in planets.

@@

In this study, we present the first spectroscopic detection of
corotating clumps of cool plasma around a CPV.  We identified
TIC~141146667 in previous work \cite{Bouma2024} by searching the TESS
two-minute data \cite{Ricker2015} for stars showing highly structured,
periodic light curves.  We selected the star for spectroscopy because
its brightness and rotation rate enable efficiently mapping out line
profile variability.  We observed it for five hours on
17~February~2024 (UT) using the High Resolution Echelle Spectrometer
(HIRES \cite{vogt_hires_1994}) on the Keck I 10m telescope.  TESS
observed the star over a four week interval from 30~January~2024 to
26~February~2024 with a duty cycle of 77\%.  During the HIRES
observations, TESS was performing a data downlink; TESS data
collection resumed 12 hours (three rotation cycles) after the spectra
were acquired.  Extended Data Figure~\ref{fig:fulllc} shows the
detailed photometric behavior of the star near the epoch of
observation; the general photometric morphology is similar before and
after the data gap.

@@

% ** keck/hires: emission beyond v/veq>1
The spectroscopy shows emission beyond the star's equatorial velocity
($v_{\rm eq}$=130\,\kms).  There are at least two distinct emission
components, separated by half a cycle in phase.  The first component
has clearer sinusoidal behaviour and is double-peaked, with peak
semi-amplitudes of $K_1$=2.1\,$v_{\rm eq}$ and 2.7\,$v_{\rm eq}$.
There is significant non-periodic variability in the emissivity of
this double-peaked component: the flux excess begins with an amplitude
of 70\% of the continuum flux early in the observation sequence, and
diminishes to 10\% by its end.  The component 180$^\circ$ opposite in
phase is detected only from $\phi$=0.2-1.0.  From $\phi$=0.2-0.5, this
latter component appears connected to the star in velocity space.
While its peak semi-amplitude of $K_1$=3.9\,$v_{\rm eq}$ is achieved
at both $\phi$=0.25 and 0.75, its amplitude similarly decreases from a
60\% excess over the continuum at the beginning of the observation
sequence to a 10\% excess by its end.  The apparent period for all
of these emission components is consistent with the photometric
\periodhr\ hour period.

@@

% * keck/hires: behavior at v/veq<1
The stellar H$\alpha$ line core, at $|\Delta v / v_{\rm
eq}|<1$, captures a superposition of emission from the star's
chromosphere, modulated by any occulting material capable of absorbing
or emitting H$\alpha$ photons.  In Figure~\ref{fig:spec}e, the data
from $\phi$=0.4-1.2 have a simple interpretation: from $\phi$=0.4-0.9,
a hot (emitting) region first gradually crosses the stellar line
profile, followed from $\phi$=0.7-1.2 by the transit of a cool
(absorbing) region. Phases $\phi$$<$0.4 show a mix of similar events,
though the time sampling is sufficiently coarse that the
interpretation is less clear.  A final exercise to quantify the
behavior in the line core is shown in Figure~\ref{fig:spec}d, where
$f_{\rm H\alpha\ core}$ denotes the summed flux at $|\Delta v / v_{\rm
eq}|<1$.  This panel shows that changes in the line core flux 
correlate with the broadband variability throughout most of the light
curve, except near $\phi$$\approx$0.5, corresponding to the transit of the
3.9\,$R_\star$ clump and the occultation of the lower-velocity clump.

@@

Our Keck/HIRES observations are the first reported time-series spectra
of a CPV, and they show that corotating circumstellar plasma clumps
are present around at least one such star.  This observation rules out
a ``starspot-only'' origin scenario for CPVs, \cite{Koen2021} because
such scenarios have no means of explaining spectroscopic emission
beyond the stellar disk.  Similarly, scenarios in which the
circumstellar material is made only of dust are also ruled out.  While
dust may be present, to explain the H$\alpha$ emission the clumps must
include plasma with a significant population of hydrogen atoms in the
$n$=3 excited state, for $n$ the principal quantum number.  More quantitatively, the observed H$\alpha$
luminosity suggests characteristic densities and masses for the
gaseous component of these clumps of $n_{\rm H} \sim
10^{11}$-$10^{12}$\,cm$^{-3}$ and $M_{\rm gas} \sim 10^{17}$\,g (see
Section~\ref{subsec:gas}).  Dust, if present, is independently
constrained to have a total mass $10^{15}\,{\rm g} < M_{\rm dust} <
10^{17}\,{\rm g}$ (see Section~\ref{subsec:dust}).  This circumstellar
material could originate either from the star or from an external
source.  Plausible external sources include an old and undetected
disk, comets, or a close-in exoplanet.
This latter scenario would make CPVs extrasolar analogs of the
Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds that
accumulate into warped plasma tori \cite{Townsend2005,Townsend2008}.
These tori tend to have dense antipodal accumulations of plasma
sculpted by tilted-dipole magnetic fields, and the transits of these
clumps produce the observed broadband photometric
variability through bound-free scattering \cite{Townsend2005} and
Thomson scattering \cite{Berry2022}.  For $\sigma$~Ori~E and almost
all of its analogs, the result is light curves that appear ``simple'',
resembling those of eclipsing binaries, and time-dynamic H$\alpha$
spectra similar to those in Figure~\ref{fig:spec}
\cite{Townsend2005,Townsend2008}.  The two known exceptions, HD~37776
and HD~64740, show complex light curves resembling CPVs
\cite{Mikulasek2020,Bouma2024} and have spectropolarimetric magnetic
field maps indicating strong contributions from higher-order magnetic
moments \cite{Kochukhov2011,Shultz2018}.  There are two implications:
first, the complexity of CPVs may be a direct consequence of magnetic
fields with highly multipolar contributions.  Second, CPVs can be a
source of astrophysical false positives in photometric classification
of eclipsing binaries and transiting exoplanets around young pre-main
sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

The material's composition -- either pure plasma, or else a dusty
plasma -- can be clarified by time-series optical and infrared
spectrophotometry.  While observations of CPVs in the optical may
suggest chromaticity consistent with dust
\cite{Tanimoto2020,Gunther2022,Koen2023}, a gray opacity source such
as electron scattering in a plasma transiting over starspots might
also produce chromatic features \cite{Rackham2018}.  The composition
and size distribution of any dust that is present could be determined
by measuring the extinction curve for one or more CPVs from
1-10\,$\mu$m.  Composition and size distributions similar to debris
from rocky bodies seen around white dwarfs \cite{Reach2009} would
indicate a rocky-body origin.  Compositions and sizes similar to the
interstellar medium could be indicative of condensed dust in an M
dwarf wind, similar to that formed in the environments of more evolved
stars \cite{Marigo2008}.

>>>>

M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water \cite{NAP26141}.
Community investment with JWST has proceeded accordingly
\cite{Redfield2024,TRAPPIST1JWSTCommunityInitiative2024}.  How an M
dwarf's evolution influences its planets---especially the retention of
their atmospheres--- has in turn become a major theme in
exoplanet and stellar astrophysics.  Previous work has established
that most M dwarfs host close-in planets \cite{Dressing2015} that 
on average are subject to long circumstellar disk lifetimes
\cite{Ribas2015}, to large doses of UV radiation \cite{France2013},
and to a high incidence of flares and coronal mass ejections
\cite{Gunther2020}.  However, despite extensive work in these areas,
the plasma and magnetospheric environments that bathe young, close-in
exoplanets remain challenging to quantify.
Understanding these environments is crucial because they directly
impact atmospheric retention and habitability of close-in planets.

@@

In this study, we present the first spectroscopic detection of
corotating clumps of cool plasma around a CPV.  We identified
TIC~141146667 in previous work \cite{Bouma2024} by searching the TESS
two-minute data \cite{Ricker2015} for stars showing highly structured,
periodic light curves.  We selected the star for spectroscopy because
its brightness and rotation rate enable measuring variability in its line
profile.  We observed it for five hours on
17~February~2024 (UT) using the High Resolution Echelle Spectrometer
(HIRES \cite{vogt_hires_1994}) on the 10\,m Keck I telescope.  TESS
observed the star over a four week interval from 30~January~2024 to
26~February~2024 with a duty cycle of 77\%.  During the HIRES
observations, TESS was performing a data downlink; TESS data
collection resumed 12 hours (three rotation cycles) after the spectra
were acquired.  Extended Data Figure~\ref{fig:fulllc} shows the
detailed photometric behavior of the star near the epoch of
observation; the general photometric morphology is similar before and
after the data gap.

@@

% ** keck/hires: emission beyond v/veq>1
The spectroscopy shows emission beyond the star's equatorial velocity
($v_{\rm eq}$=130\,\kms).  There are at least two distinct emission
components, separated by half a cycle in phase.  The first component
has clearer sinusoidal behaviour and is double-peaked, with peak
semi-amplitudes of $K_1$=2.1\,$v_{\rm eq}$ and 2.7\,$v_{\rm eq}$.
There is significant non-periodic variability in the emissivity of
this double-peaked component: the flux excess begins with an amplitude
of 70\% above the continuum early in the observation sequence, and
diminishes to 10\% by its end.  The component 180$^\circ$ opposite in
phase is detected only from $\phi$=0.2-1.0.  From $\phi$=0.2-0.5, this
latter component appears connected to the star in velocity space.
While its peak semi-amplitude of $K_1$=3.9\,$v_{\rm eq}$ is achieved
at both $\phi$=0.25 and 0.75, its amplitude similarly decreases from a
60\% excess over the continuum at the beginning of the observation
sequence to a 10\% excess by its end.  The apparent period for all
of these emission components is consistent with the photometric
\periodhr\ hour period.

@@

% * keck/hires: behavior at v/veq<1
The H$\alpha$ line core at $|\Delta v / v_{\rm
eq}|<1$ originates from a superposition of emission from the star's
chromosphere, modulated by occulting material capable of absorbing
or emitting H$\alpha$ photons.  In Figure~\ref{fig:spec}e, the data
from $\phi$=0.4-1.2 have a simple interpretation: from $\phi$=0.4-0.9,
a hot (emitting) region first gradually crosses the stellar line
profile, followed from $\phi$=0.7-1.2 by the transit of a cool
(absorbing) region. Phases $\phi$$<$0.4 show a mix of similar events,
though the time sampling is sufficiently coarse that the
interpretation is less clear.  A final exercise to quantify the
behavior in the line core is shown in Figure~\ref{fig:spec}d, where
$f_{\rm H\alpha\ core}$ denotes the summed flux at $|\Delta v / v_{\rm
eq}|<1$.  This panel shows that changes in the line core flux 
correlate with the broadband variability throughout most of the light
curve, except near $\phi$$\approx$0.5, corresponding to the transit of the
3.9\,$R_\star$ clump and the occultation of the lower-velocity clump.

@@

Our Keck/HIRES observations are the first reported time-series spectra
of a CPV, and they show that corotating circumstellar plasma clumps
are present around at least one such star.  This observation rules out
a ``starspot-only'' origin for the CPVs, \cite{Koen2021} because
such scenarios have no means of explaining spectroscopic emission
beyond the stellar disk.  Similarly, scenarios in which the
circumstellar material is made only of dust are also ruled out.  While
dust may be present, to explain the H$\alpha$ emission the clumps must
include plasma with a significant population of hydrogen atoms in the
$n$=3 excited state, where $n$ denotes the principal quantum number.
More quantitatively, the observed H$\alpha$ luminosity suggests
characteristic densities and masses for the gaseous component of these
clumps of $n_{\rm H} \sim 10^{11}$-$10^{12}$\,cm$^{-3}$ and $M_{\rm
gas} \sim 10^{17}$\,g (see Section~\ref{subsec:gas}).  Dust, if
present, is independently constrained to have a total mass
$10^{15}\,{\rm g} < M_{\rm dust} < 10^{17}\,{\rm g}$ (see
Section~\ref{subsec:dust}).  This circumstellar material could
originate either from the star or from an external source.  Plausible
external sources include an old and undetected disk, comets, or a
close-in exoplanet.  This latter scenario would make CPVs extrasolar
analogs of the Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds that
accumulate into warped plasma tori \cite{Townsend2005,Townsend2008}.
These tori tend to have dense antipodal accumulations of plasma
sculpted by tilted-dipole magnetic fields, and the transits of these
clumps produce broadband photometric
variability through bound-free scattering \cite{Townsend2005} and
Thomson scattering \cite{Berry2022}.  For $\sigma$~Ori~E and almost
all of its analogs, the result is light curves that appear ``simple'',
resembling those of eclipsing binaries, and time-dynamic H$\alpha$
spectra similar to those in Figure~\ref{fig:spec}
\cite{Townsend2005,Townsend2008}.  The two known exceptions, HD~37776
and HD~64740, show complex light curves resembling CPVs
\cite{Mikulasek2020,Bouma2024} and have spectropolarimetric magnetic
field maps indicating strong contributions from higher-order magnetic
moments \cite{Kochukhov2011,Shultz2018}.  There are two implications:
first, the complexity of CPVs may be a direct consequence of magnetic
fields with highly multipolar contributions.  Second, CPVs can be a
source of astrophysical false positives in photometric classification
of eclipsing binaries and transiting exoplanets around young pre-main
sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

The material's composition -- either pure plasma, or else a dusty
plasma -- can be clarified by time-series optical and infrared
spectrophotometry.  While observations of CPVs in the optical 
suggest a chromaticity consistent with dust
\cite{Tanimoto2020,Gunther2022,Koen2023}, a gray opacity source such
as electron scattering in a plasma transiting over starspots might
also produce chromatic features \cite{Rackham2018}.  The composition
and size distribution of any dust that is present could be determined
by measuring the extinction curve for one or more CPVs from
1-20\,$\mu$m.  Composition and size distributions similar to debris
from rocky bodies seen around white dwarfs \cite{Reach2009} would
indicate a rocky-body origin.  Compositions and sizes similar to the
interstellar medium could be indicative of condensed dust in an M
dwarf wind, similar to that formed in the environments of more evolved
stars \cite{Marigo2008}.
